---
title: "Gingival results"
author: "Samuel Gamboa-Tuz, Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    html_document:
        toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(microbiomeMarker)
library(benchdamic)
library(MicrobiomeBenchmarkData)
library(ggpubr)
```

## Import data

Create a temporary directory to store the files:
```{r}
fName <- system.file(
    "extdata", "gingival_results_2024-09-18.zip",
    package = "MicrobiomeBenchmarkDataLefse", mustWork = TRUE
)
tmpDir <- tempfile()
unzip(zipfile = fName, exdir = tmpDir)
fls <- list.files(tmpDir, full.names = TRUE)
names(fls) <- list.files(tmpDir, full.names = FALSE)
```

Read the data and make it uniform:

```{r}
## Results from LEfSe (run in Docker)
lefseDat <- read_tsv(
    grep("\\.res", fls, value = TRUE),
    col_names = c("feature", "x", "sample", "lda", "rawP"),
    show_col_types = FALSE
) |> 
    select(-x) |> 
    drop_na() |> 
    mutate(
        feature = sub("(OTU)(97)(\\d+)$", "\\1_\\2.\\3", feature)
    ) |> 
    mutate(rawP = as.double(rawP)) |> 
    mutate(adjP = p.adjust(rawP)) |> 
    dplyr::rename(enGrp = sample) |> 
    mutate(software = "LEfSe") |> 
    tibble::column_to_rownames(var = "feature") |> 
    mutate(
        lda = case_when(
            enGrp == "subgingival_plaque" ~ lda,
            enGrp == "supragingival_plaque" ~ -lda,
        )
    ) |> 
    arrange(lda)
    
## Results from lefser
lefserDat <- read_csv(
    grep("lefser\\.csv", fls, value = TRUE),
    show_col_types = FALSE
) |> 
    tibble::column_to_rownames(var = "features") |> 
    dplyr::rename(lda = scores) |> 
    mutate(
        ## I got the enriched groups by comparing this result with
        ## lefseDat and mmDat
        enGrp = case_when(
            lda < 0 ~ "supragingival_plaque",
            lda > 0 ~ "subgingival_plaque",
        ),
        # lda = abs(lda),
        rawP = 0.0099,
        adjP = 0.099
    ) |> 
    mutate(software = "lefser") |> 
    arrange(lda)

## Results from microbiomeMarker
mmDat <- readRDS(
    grep("mm\\.rds", fls, value = TRUE),
) |>
    marker_table() |> 
    as.matrix() |> 
    as.data.frame() |> 
    magrittr::set_rownames(NULL) |> 
    tibble::column_to_rownames("feature") |> 
    dplyr::rename(
        enGrp = enrich_group,
        lda = ef_lda,
        rawP = pvalue,
        adjP = padj
    ) |> 
    mutate_at(
        .vars = c("lda", "rawP", "adjP"),
        .funs = as.double
    ) |> 
    mutate(software = "microbiomeMarker") |> 
    mutate(
        lda = case_when(
            enGrp == "subgingival_plaque" ~ lda,
            enGrp == "supragingival_plaque" ~ -lda,
        )
    ) |> 
    arrange(lda)

DA <- list(
    LEfSe = lefseDat,
    lefser = lefserDat,
    microbiomeMarker = mmDat
)
map_int(DA, nrow)
```
## Features detected in all methods

```{r}
x <- intersect(
    intersect(rownames(mmDat), rownames(lefserDat)),
    rownames(lefseDat)
)
length(x)
```

## Comparison of absolute LDA scores

```{r}
DA |> 
    bind_rows() |>
    ggplot(aes(lda)) +
    geom_histogram(aes(fill = software), binwidth = 0.1) +
    scale_fill_viridis_d(option = "D", name = "Software") +
    labs(
        x = "LDA score",
        y = "Frequency"
    ) +
    theme_minimal()
```
```{r}
DA |> 
    bind_rows() |> 
    mutate(
        sig = case_when(
            rawP < 0.05 ~ TRUE,
            TRUE ~ FALSE
        )
    ) |> 
    pull(sig) |> 
    table()
```

## Benchdamic-like plots

Create a list object compatible with the `benchdamic` workflow:

```{r}
createBenchObj <- function(x) {
    pValMat <- x[, c("rawP", "adjP"), drop = FALSE]
    statInfo <- x
    name = unique(x[["software"]])
    list(
        pValMat = pValMat,
        statInfo = statInfo,
        name =  name
    )
}
DA_output <- map(DA, createBenchObj)
str(DA_output, max.level = 2)
```
Get biological information

```{r, message=FALSE, warning=FALSE}
tse <- getBenchmarkData("HMP_2012_16S_gingival_V35", dryrun = FALSE)[[1]]
bioData <- tse |> 
    rowData() |> 
    as.data.frame()
```

### Plot of difference of putatitive TRUE and FALSE positives

```{r}
cbfColors <- c("#D55E00", "#009E73", "#0072B2")
posData <- createPositives(
    object = DA_output,
    priorKnowledge = bioData, enrichmentCol = "taxon_annotation",
    namesCol = NULL,
    slot = "statInfo", colName = "lda", type = "logfc",
    direction = "lda",
    threshold_pvalue = 1, threshold_logfc = 0,
    top = seq(5, 100, 5),
    alternative = "g",
    verbose = FALSE,
    TP = list(
        c("UP Abundant", "anaerobic"),
        c("DOWN Abundant", "aerobic")
    ),
    FP = list(
        c("UP Abundant", "aerobic"),
        c("DOWN Abundant", "anaerobic")
    )
)
```

```{r}
posPlot <- plotPositives(posData)
for (i in seq_along(posPlot$layers)) {
    ## Remove layers (geom_line and geom_point)
    ## This is necessary to add new layers with shapes and colors of choice
    posPlot$layers[[1]] <- NULL
}
posPlot <- posPlot + 
    geom_line(
        aes(x = top, y = TP - FP, linetype = method)
    ) +
    geom_point(
        aes(x = top, y = TP - FP, shape = method)
    ) +
    scale_linetype(name = "Software") +
    scale_shape_manual(values = c(1, 2, 3), name = "Software") +
    scale_color_manual(values = cbfColors, name = "Software") +
    theme_minimal() +
    labs(title = NULL, subtitle = NULL) +
    ylab("Putative TP - Putative FP") +
    xlab("# of top features") +
    theme(
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)   
    )
posPlot
```
### Enrichemnt plot

```{r}
enData <- createEnrichment(
    object = DA_output,
    priorKnowledge = bioData, enrichmentCol = "taxon_annotation",
    namesCol = NULL,
    slot = "statInfo", colName = "lda", type = "logfc",
    direction = "lda",
    ## No threshold since all results included here
    ## are significant
    threshold_pvalue = 1, threshold_logfc = 0,
    top = NULL,
    alternative = "greater",
    verbose = FALSE
)
enPlot <- plotEnrichment(enData, "taxon_annotation", c("aerobic", "anaerobic"))
enPlot <- enPlot +
    labs(x = "Software") +
    theme_minimal()
enPlot
```


```{r}
myFun <- function(x) {
    x |> 
        tibble::rownames_to_column(var = "direction")
}
enDatMod <- map(enData, ~ {
    .x$summaries |> 
        map(myFun) |> 
        bind_rows() |> 
        pivot_longer(
            names_to = "annotation", values_to = "ntaxa",
            cols = c(aerobic, anaerobic, facultative_anaerobic)
        ) |> 
        filter(!is.na(ntaxa))
}) |> 
    bind_rows(.id = "software") |> 
    mutate(
        direction = case_when(
            direction == "DOWN Abundant" ~ "Supragingival",
            direction == "UP Abundant" ~ "Subgingival"
        )
    ) |>  
    mutate(
        sig = case_when(
            pvalue < 0.05 & pvalue >= 0.01 ~ "*" ,
            pvalue < 0.01 & pvalue >= 0.001 ~ "**",
            pvalue < 0.001 ~ "***",
            TRUE ~  ""
        )
    )
```

```{r}
enPlotMod <- enDatMod |> 
    ggplot(aes(software, ntaxa)) +
    geom_col(
        aes(fill = annotation),
        position = position_dodge2(width = 0.9)
    ) +
    geom_text(
        aes(label = sig, color = annotation), 
        position = position_dodge2(width = 0.9)
    ) +
    facet_wrap(~direction) +
    scale_fill_manual(
        values = cbfColors, name = "Annotations"
    ) +
    labs(
        x = "", y = "Number of DA taxa"
    ) +
    guides(color="none") +
    theme_minimal() +
    theme(
        axis.title = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10)   
    )
enPlotMod
```

### Combined plot

```{r, fig.width=12, fig.height=5}
pts <- ggarrange(
    enPlotMod, posPlot,
    nrow = 1, labels = c("a)", "b)"),
    widths = c(1, 1)
)
pts
```

## Session information

```{r}
sessioninfo::session_info()
```






